Indel Historian: fast and accurate ancestral sequence reconstruction using weighted finite-state transducers

### 2016-07-22 doc/main.tex
writing

The study of sequence evolution and molecular phylogenetics gives rise to
a different set of criteria for measuring alignment accuracy.
As an example, for the purposes of predicting a protein structure,
functionally equivalent sequences in a rapidly-evolving region (such as a loop)
may be considered alignable even if they have been entirely replaced by a series of indel mutations
(e.g. the sequences $X$ and $Y$ in the mutation trajectory $AXB \to AXYB \to AYB$),
but for the purposes of evolutionary analysis, one generally wants to count every indel event that has occurred.
Previous work showed that, for the purposes of estimating
molecular evolutionary parameters---such as the rates of insertion-deletion events \citep{Westesson2012-zg},
the ratio between synonymous and nonsynonymous substitution rates \citep{Redelings2014},
or other evolutionary analyses \citep{LoytynojaGoldman2008}---it is best to use an explicit probabilistic model of the sequence evolution process
and to treat sequence alignment rigorously as a missing data estimation problem
under this model.

@@

Our earlier benchmark of alignment methods on simulated sequence data
showed that most alignment algorithms introduce systematic biases into the
estimation of indel rates, and that these biases are typically substantially worse
at higher indel rates \citep{Westesson2012-zg}.
ProtPal was found to be the most accurate algorithm for indel rate estimation,
followed by PRANK and then MUSCLE.
However, the implementation of ProtPal published with that benchmark was too slow
for practical use.
Furthermore, no benchmark of ProtPal on structural alignment benchmarks was available,
and---although we have noted that structure prediction is a different task---it would be
useful to know just how different it is, according to the metrics.

>>>>

The study of sequence evolution and molecular phylogenetics gives rise to
a different set of criteria for measuring alignment accuracy.
Several independent studies have showed that, for the purposes of estimating
molecular evolutionary parameters---such as the rates of insertion-deletion events \citep{Westesson2012-zg},
the ratio between synonymous and nonsynonymous substitution rates \citep{Redelings2014},
or other evolutionary analyses \citep{LoytynojaGoldman2008}---it is best to use an explicit probabilistic model of the sequence evolution process
and to treat sequence alignment rigorously as a missing data estimation problem
under this model.
One conceptual explanation for this phenomenon is that, for the purposes of predicting a protein structure,
functionally equivalent sequences in a rapidly-evolving region (such as a loop)
may be considered alignable even if they have been entirely replaced by a series of indel mutations
(e.g. the sequences $X$ and $Y$ in the mutation trajectory $AXB \to AXYB \to AYB$),
but for the purposes of evolutionary analysis, one generally wants to count every indel event that has occurred.

@@

Our earlier benchmark of alignment methods on simulated sequence data
showed that most alignment algorithms introduce systematic biases into the
estimation of indel rates, and that these biases are typically substantially worse
at higher indel rates \citep{Westesson2012-zg}.
ProtPal was found to be the most accurate algorithm for indel rate estimation,
followed by PRANK and then MUSCLE.
However, the implementation of ProtPal published with that benchmark was too slow
for practical use.
Furthermore, no benchmark of ProtPal on structural alignment benchmarks was available,
and---although we have noted that structure prediction is a different task from evolutionary analysis---it would be
useful to know just how different it is, according to the metrics.
